---
title: "Data Analysis and Visualisation"
subtitle: "Part 2"
author: "Thomas Wise"
format: 
  revealjs:
    smaller: true
    theme: dark
editor: visual
---

## Session Objectives 

- Understand and apply advanced structural and design elements to ggplot2 visualisations 
- Be able to create complex visualisations in line with Welsh Government design requirements 
- Understand how to build simple machine learning models (particularly CART Models) 
- Apply and develop simple machine learning models effectively. 

## Today's Plan 

- Section 1: Design elements of ggplot2 visualisations 
- Exercise 1: Applying design elements to ggplot2 
- Break 
- Section 2: Developing Machine Learning Models 
- Exercise 2: Applying Machine Learning Models 

# But first - Setup! 

# Design elements of ggplot2 visualisations 

## Specifying Axis's (recap)

Simple short cuts:

-   `lims(x = c(min, max), y = c(min, max))`

Advanced used:

-   `scale_x_continuous()` / `scale_y_continuous()`
-   `scale_x_sqrt()` / `scale_y_sqrt()`
-   `scale_x_log10()` / `scale_y_log10()`

With these having the internal arguments:

-   breaks, minor_breaks, n.breaks
-   labels, limits

## Specifying Labels (recap)

You can simply specify axis labels using the `labs()` function with the internal arguments:

-   `title =` for the plot title
-   `subtitle =` for the plot subtitle
-   `caption =` for the plot caption
-   `x` / `y` for x or y axises respectively

Advanced labels can be given through:

-   `annotate()` function combined with geom `text` or `rect`
-   `geom_vline()` for a vertical line
-   `geom_hline()` for a horizontal line

## Basic Colours in ggplot2

Adding colours to a visualization can be done either through

-   Key Names: *"red"*, *"blue"*
-   Hexcodes/RGB Specification: *"#FF0000"*, *"#0000FF"*

And can be specified in several places in the ggplot code:

-   `colour =` - for points, or edge/line colours
-   `fill =` - for internal colours or fill colours

## Specifying Palettes using RColourBrewer

The CRAN package `RColourBrewer`, adds many ready-to-use discrete variable palettes such as:

-   Colour blind friendly palettes
-   Specific colour schemes
-   Convergent or divergent schemes

These can be specified in your plot using:

-   `+ scale_colour_brewer(palette = "XX")` - for colour cases
-   `+ scale_fill_brewer(palette = "XX")` - for fill cases

To view all available palettes from RColourBrewer use:

-   `display.brewer.all()`, specifying `colorblindFriendly = TRUE` accordingly

## Types of RColourBrewer Palette

RColourBrewer specifies three types of palette:

-   Sequential palettes: suited to ordered data which progress from low to high (gradient) these include: *Blues, BuGn, BuPu, GnBu, Greens, Greys, Oranges, OrRd, PuBu, PuBuGn, PuRd, Purples, RdPu, Reds, YlGn, YlGnBu YlOrBr, YlOrRd*
-   Qualitative palettes: suited to represent nominal or categorical data, without indicating magnitude differences between groups: *Accent, Dark2, Paired, Pastel1, Pastel2, Set1, Set2, Set3*
-   Diverging palettes: suited for putting an emphasis on mid-range critical values, and extremes at both ends of the data range: : *BrBG, PiYG, PRGn, PuOr, RdBu, RdGy, RdYlBu, RdYlGn, Spectral*

## Continuous Variables and Colour Palettes

If you are using continuous variables, such as height, weight or size, use:

-   `scale_colour_gradient()` - specifying the `low` and `high` colours accordingly
-   `scale_fill_gradient()` - specifying the `low` and `high` colours accordingly

## Linking Colours to Variables

To link colours to variables simply, you can use the function:

-   `scale_color_manual(values = c(VAR = COL))`

This manual specification of colour should then be paired with specifying the colour variable inside the `mapping = aes(colour = VAR)`.

This can be extended through assigning this to a variable!

## Applying colour accessibility guidelines

Colour accessibility guidelines are provided by the [Government Analysis Function](https://analysisfunction.civilservice.gov.uk/policy-store/data-visualisation-colours-in-charts/). And highlights the following:

-   Use colour consistently
-   Consider colour associations
-   Keep backgrounds and layouts plain
-   Consider low vision and colour blindness
-   Consider what the plot would look like in grey scale

## Applying colour accessibility guidelines

The GSS Colour Palette:

-   Dark Blue, #12436D, (18, 67, 109)RGB, (83, 39, 0, 57)CYMK
-   Turquoise, #28A197, (40, 161, 151)RGB, (75, 0, 6, 37)CYMK
-   Dark Pink, #801650, (128, 22, 80)RGB, (0, 83, 38, 50)CYMK
-   Orange, #F46A25, (244, 106, 37)RGB, (0, 57, 85, 4)CYMK

All these colours have a contrast ratio of a least 3 to 1 against white

Additional Colours: Dark Grey #12436D, Light Purple #A285D1

## Using preset themes

-   `theme_bw()`
-   `theme_classic()`
-   `theme_dark()`
-   `theme_light()`
-   `theme_minimal()`
-   `theme_void()`

## Using specifying manual themes

By using the `theme()` function as a foundation, you can specify almost every element of a visualisation plot.

But remember to call these functions when using specific elements

-   Line elements: `element_line()`
-   Rectangular/area elements: `element_rect()`
-   Text elements: `element_text()`

You can use these to rotate axis labels too! 
For example: 

- `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))`

## Specifying Layouts using faceting

The `facet_grid()` and `facet_wrap()` functions can be used to segment the visualisation by a variable.

This requires the specification of variable using `vars()`.

```{r}
#| echo: TRUE
#| eval: TRUE

library(ggplot2)

# Example comparing diamond price to carat in ggplot
ggplot(data = diamonds) + 
  geom_point(mapping = aes(y = price, x = carat, colour = color)) + 
  facet_grid(cols = vars(color))
```

## Specifying Layouts using ggarrange

The `ggarrange()` function can be used to combine ggplot visualisations into a single plot.

This requires the variables:

-   ncol /nrow - numeric values
-   ..., stored ggplot graphics
-   common.legend = TRUE/FALSE

This can help compress visualisations into a single plot and make them more interpretable.

# Welsh Government Data Visualisation Theme

# Exercise 1: Applying design elements to ggplot2

# Break

# Developing Machine Learning Models 

## Note:

As always, there are multiple packages which can be used to apply Machine Learning models. 

Today we will focus on using `tidymodels`, which use a tidyverse approach.

However, other packages such as `e1071` or `rpart` are available for specific uses

## When to apply machine learning

Machine Learning problems can be seen as being part of three main groups: 

- To predict
- To classify 
- To cluster, group or investigate 

These each link to a different type of Machine Learning Method. 

## Types of Machine Learning 

Supervised Learning: Those which use *tagged* or *labelled* data to learn

- Regression: 
  - Problem aim: Prediction 
  - Outcome Variable: Continuous 
  
- Classification: 
  - Problem aim: Classify 
  - Outcome variable: Categorical 
  
Unsupervised Learning: Those which use *untagged* or *unlabeled* data to learn 

- Clustering:
  - Problem aim: Grouping or Clustering 
  - Outcome variable: No specific variable type
  
## Matching variables to the methodology

When considering machine learning, the methodology will typically depend on the data available

- Classification methods can only be used on Categorical outcome variables 
- Prediction methods can only be used on Continuous outcome variables 

Of course, there is nuance in this, especially when you can convert continuous to categorical variables. 

## Why Tidymodels 

Tidymodels looks to apply the tidyverse style and readability to create a unified approach to building models. Taking advantage of the multitude of different packages each requiring different syntaxes and approaches. 

This allows for easy scaling of models to include underwritters like Spark

## Building Models 

To build effective models, you should take your data, and split it into: 

- **Training data** - This is where we develop our model [Typically 80% of the data]
- **Testing data** - This is where we test the models accuracy [Typically 20% of the data]
- **Validation data** (optional) - This is how we validate our findings

As this helps to reduce the risk of over fitting and improves models accuracy. 

## How to split data 

When splitting data, we typically want a 80:20 Training:Test split. 

This can be achieved with tidymodels through the `initial_split()` function 

- `data_split <- initial_split(data, prop = 0.8)`
- `data_training <- training(data_split)`
- `data_test <- testing(data_split)`

To split for validation you can use:

- `data_split <- initial_split(data, prop = c(0.6, 0.2))`
- `data_valid <- validation(data_split)`

## Introducing Reproducibility 

- Being able to reproduce the splits you create in your data is vital.
- You can control the randomness found, using `set.seed()` function

- We will cover more on embedding reproducibility next week. 

## Building a tidy model

Importantly, when using tidy models, we are required to *set an engine*, defining how the model should be developed for example: 

```{r}
#| echo: TRUE
#| eval: false

# Linear regression model 
lm_model <- 
  liner_reg() |> 
  set_engine("lm") 

# Classification Decision tree 
c_dt_model <- 
  decision_tree() |> 
  set_engine("rpart") |> 
  set_mode("classification")
```

In these cases you can also add the `translate()` function to convert the produced output into the same format as the traditional engine. 

## Some Common & Accessible Machine Learning Methods 

- Supervised Learning 
  - Regression: 
    - Linear Regressions / GLM - `linear_reg()` with engines: `"lm"`
    - Decision Trees - `decision_tree()` with engines: `"rpart"` and mode `"regression"`
    - Support Vector Regression - `svm_linear()` with engines: `"kernlab"` and mode `"regression"`
    
    
## Some Common & Accessible Machine Learning Methods 
- Supervised Learning
  - Classification: 
    - Logistic Regressions - `logistic_reg()` with engines: `"glm"`
    - Decision Trees - `decision_tree()` with engines: `"rpart"` and mode `"classification"`
    - Support Vector Machines - `svm_linear()` with engines: `"kernlab"` and mode `"classification"`

## Fitting Models 

Once your model has been created, you can utilise the `fit()` function 

```{r}
#| echo: TRUE
#| eval: false

# Linear regression model 
lm_model <- 
  liner_reg() |> 
  set_engine("lm") 

lm_model_fit <- 
  lm_model |> 
  fit(data = data, 
      y ~ x)

```

You can then view the outcomes using: 

```{r}
#| echo: TRUE
#| eval: false

model_res <- 
  lm_model_fit |>
  extract_fit_engine() |>
  summary()

```

## Making Predictions

You can then use this model, to make predictions using: 

```{r}
#| echo: TRUE
#| eval: false

# Linear regression model 
lm_model_fit <- 
  lm_model |> 
  fit(data = data, 
      y ~ x)

# Model Prediction 
lm_model_prediction <- 
  predict(lm_model_fit, 
          new_data = data_new)

# Bind the prediction with the "truth"
model_res <- bind_cols(lm_model_prediction, 
                       data_new |> select(x))

```

## Evaluating Models 

This will create predictions based on the data provided, which can then be used to evaluate model effectiveness 

```{r}
#| echo: TRUE
#| eval: false

# Set Model Effectiveness Measures (for linear regression)

model_metrics <- metric_set(rmse, rsq, mae)

# Apply the evaluation methods 

model_metrics(model_res, truth = x, estimate = .pred)

```

Please note, the model metrics, will vary depending on the method being used! 

# Exercise 2: Applying Machine Learning Models 

## Next Session: 

- Reproducibility 
- 16th June, 12:30-14:00

## Resources: 

- [Tidy Modeling with R](<https://www.tmwr.org/>)
- [Tidy Modeling Cheatsheet](<https://www.tidymodels.org/find/parsnip/>)
- [Learning Statistics with R - Danielle Navarro](https://learningstatisticswithr.com/lsr-0.6.pdf)
- [Introduction to Statistical Learning - James, Witten, Hastie & Tibshirani](https://www.statlearning.com/)





