---
title: "Data Analysis and Visualisation"
subtitle: "Part 2"
author: "Thomas Wise"
format: 
  revealjs:
    smaller: true
    theme: dark
editor: visual
---

## Session Objectives 

- Understand and apply advanced structural and design elements to ggplot2 visualisations 
- Be able to create complex visualisations in line with Welsh Government design requirements 
- Understand how to build simple machine learning models (particularly CART Models) 
- Apply and develop simple machine learning models effectively. 

## Today's Plan 

- Section 1: Design elements of ggplot2 visualisations 
- Exercise 1: Applying design elements to ggplot2 
- Break 
- Section 2: Developing Machine Learning Models 
- Exercise 2: Applying Machine Learning Models 

## 

# Design elements of ggplot2 visualisations 

# Exercise 1: Applying design elements to ggplot2

# Break

# Developing Machine Learning Models 

## Note:

As always, there are multiple packages which can be used to apply Machine Learning models. 

Today we will focus on using `tidymodels`, which use a tidyverse approach.

However, other packages such as `e1071` or `rpart` are available for specific uses

## When to apply machine learning

Machine Learning problems can be seen as being part of three main groups: 

- To predict
- To classify 
- To cluster, group or investigate 

These each link to a different type of Machine Learning Method. 

## Types of Machine Learning 

Supervised Learning: Those which use *tagged* or *labelled* data to learn

- Regression: 
  - Problem aim: Prediction 
  - Outcome Variable: Continuous 
  
- Classification: 
  - Problem aim: Classify 
  - Outcome variable: Categorical 
  
Unsupervised Learning: Those which use *untagged* or *unlabeled* data to learn 

- Clustering:
  - Problem aim: Grouping or Clustering 
  - Outcome variable: No specific variable type
  
## Matching variables to the methodology

When considering machine learning, the methodology will typically depend on the data available

- Classification methods can only be used on Categorical outcome variables 
- Prediction methods can only be used on Continuous outcome variables 

Of course, there is nuance in this, especially when you can convert continuous to categorical variables. 

## Why Tidymodels 

Tidymodels looks to apply the tidyverse style and readability to create a unified approach to building models. Taking advantage of the multitude of different packages each requiring different syntaxes and approaches. 

This allows for easy scaling of models to include underwritters like Spark

## Building Models 

To build effective models, you should take your data, and split it into: 

- **Training data** - This is where we develop our model [Typically 80% of the data]
- **Testing data** - This is where we test the models accuracy [Typically 20% of the data]
- **Validation data** (optional) - This is how we validate our findings

As this helps to reduce the risk of over fitting and improves models accuracy. 

## How to split data 

When splitting data, we typically want a 80:20 Training:Test split. 

This can be achieved with tidymodels through the `initial_split()` function 

- `data_split <- initial_split(data, prop = 0.8)`
- `data_training <- training(data_split)`
- `data_test <- testing(data_split)`

To split for validation you can use:

- `data_split <- initial_split(data, prop = c(0.6, 0.2))`
- `data_valid <- validation(data_split)`

## Introducing Reproducibility 

- Being able to reproduce the splits you create in your data is vital.
- You can control the randomness found, using `set.seed()` function

- We will cover more on embedding reproducibility next week. 

## Building a tidy model

Importantly, when using tidy models, we are required to *set an engine*, defining how the model should be developed for example: 

```{r}
#| echo: TRUE
#| eval: false

# Linear regression model 
lm_model <- 
  liner_reg() |> 
  set_engine("lm") 

# Classification Decision tree 
c_dt_model <- 
  decision_tree() |> 
  set_engine("rpart") |> 
  set_mode("classification")
```

In these cases you can also add the `translate()` function to convert the produced output into the same format as the traditional engine. 

## Some Common & Accessible Machine Learning Methods 

- Supervised Learning 
  - Regression: 
    - Linear Regressions / GLM - `linear_reg()` with engines: `"lm"`
    - Decision Trees - `decision_tree()` with engines: `"rpart"` and mode `"regression"`
    - Support Vector Regression - `svm_linear()` with engines: `"kernlab"` and mode `"regression"`
    
    
## Some Common & Accessible Machine Learning Methods 
- Supervised Learning
  - Classification: 
    - Logistic Regressions - `logistic_reg()` with engines: `"glm"`
    - Decision Trees - `decision_tree()` with engines: `"rpart"` and mode `"classification"`
    - Support Vector Machines - `svm_linear()` with engines: `"kernlab"` and mode `"classification"`

## Fitting Models 

Once your model has been created, you can utilise the `fit()` function 

```{r}
#| echo: TRUE
#| eval: false

# Linear regression model 
lm_model <- 
  liner_reg() |> 
  set_engine("lm") 

lm_model_fit <- 
  lm_model |> 
  fit(data = data, 
      y ~ x)

```

You can then view the outcomes using: 

```{r}
#| echo: TRUE
#| eval: false

model_res <- 
  lm_model_fit |>
  extract_fit_engine() |>
  summary()

```

## Making Predictions

You can then use this model, to make predictions using: 

```{r}
#| echo: TRUE
#| eval: false

# Linear regression model 
lm_model_fit <- 
  lm_model |> 
  fit(data = data, 
      y ~ x)

# Model Prediction 
lm_model_prediction <- 
  predict(lm_model_fit, 
          new_data = data_new)

# Bind the prediction with the "truth"
model_res <- bind_cols(lm_model_prediction, 
                       data_new |> select(x))

```

## Evaluating Models 

This will create predictions based on the data provided, which can then be used to evaluate model effectiveness 

```{r}
#| echo: TRUE
#| eval: false

# Set Model Effectiveness Measures (for linear regression)

model_metrics <- metric_set(rmse, rsq, mae)

# Apply the evaluation methods 

model_metrics(model_res, truth = x, estimate = .pred)

```

Please note, the model metrics, will vary depending on the method being used! 

# Exercise 2: Applying Machine Learning Models 

## Next Session: 

- Reproducibility 
- 16th June, 12:30-14:00

## Resources: 

- [Tidy Modeling with R](<https://www.tmwr.org/>)
- [Tidy Modeling Cheatsheet](<https://www.tidymodels.org/find/parsnip/>)
- [Learning Statistics with R - Danielle Navarro](https://learningstatisticswithr.com/lsr-0.6.pdf)
- [Introduction to Statistical Learning - James, Witten, Hastie & Tibshirani](https://www.statlearning.com/)





